The ppmData R-package for setting up spatial point process models

### 2022-09-08 paper/paper.Rmd
update

Spatial data, the locations of objects in a spatial region, is a common type of data recorded in ecology, environmental sciences and epidemiology amongst many others. An appropriate statistical approach for such data are spatial point process models [e.g. @cressie_statistics_1993; @baddeley_practical_2000]. Spatial point processes can be used to understand the location of objects to inform decision making and improve ecological, health and other socio-economic outcomes. For example, the locations of individuals from a species' population are typically not distributed completely at random, but rather are observed at locations with favorable habitat conditions for that species. We might expect these species sightings are distributed according to relevant covariates.

Inhomogeneous Poisson Process Models (IPPM) allow for the spatial intensity of a point process to vary across space, usually as a function of spatially referenced covariates. One challenge when fitting an Inhomogeneous Poisson Process Model is the integral used to described the intensity surface in the log-likelihood (as described below) has no known closed analytic solution. For IPPMs, the Berman-Turner device [@berman_approximating_1992] is commonly used to approximate the model using quadrature within a weighted Poisson generalized linear model. The type of quadrature scheme can have important ramifications for overall model fitting and inference [@warton_poisson_2010]. Approximating the integral accurately often comes at an increased computation cost, where a large number of quadrature points are required to robustly estimate the integral [@renner_point_2015].

@@

Here we present the `ppmData` R package that is designed to setup a quadrature scheme using Dirichlet tessellation for fitting spatial point process models. Our initial intention was to develop multiple species (marked) quadrature in order to improve computation of multiple species Poisson point process models. Since then, the package has involved into one which can be useful for fitting univariate response inhomogeneous spatial point process, as discussed in this paper. We describe a how to set up a inhomogeneous Poisson point process quadrature and provide a simple demonstration on how one can fit an inhomogeneous Poisson point process using a `ppmData` object.

@@

An Inhomogeneous Poisson Process Models is a statistical model used to describe a random point pattern for $n$ observed locations, ($\bm{y}=(y(\bm{s}_1),y(\bm{s_2}),\ldots,y(\bm{s}_n))$, in a spatial window $\mathcal{A}\subset R^2$. An Inhomogeneous Poisson Process Model assumes that the number of points in $\mathcal{A}$, $n$, is a Poisson random variable with total intensity $\Lambda$ defined as $\Lambda = \int_{\bm{s}\in\mathcal{A}}\lambda(\bm{s})d\bm{s}$. The spatial function $\lambda(\bm{s})$ is usually a non-constant function and defines the spatial surface that the spatial locations $\bm{y}$ are independently drawn from [e.g. @cressie_statistics_1993]. For convenience, we use subscript notation for particular spatial locations, (for example) $\lambda_i=\lambda(\bm{s}_i)$ is the intensity surface of the point process at the $i^{th}$ location with coordinates $\bm{s}_i$. The intensity surface can be defined as $\log\{\lambda_i\}=\bm{x}_i^{\top}\bm{\beta}$ where $\bm{x}_i$ is a vector of spatially located covariates and $\bm{\beta}$ is a corresponding vector of coefficients.

@@

```{r fig1, fig.width=6, fig.height=3,fig.align='center',cache=TRUE, warning=FALSE, echo=FALSE, message=FALSE,fig.cap = "Quadrature schemes generated using the ppmData package for the species \\textit{Tasmaphena sinclairi} using Tasmania, Australia, as our spatial region. Red points are the known presences of \\textit{Tasmaphena sinclairi}. Black points are the background quadrature locations. A) Quasi-random quadrature where the integration points are generated using a quasi-random areal sample. Weights for each integration point are calculated as the dual of the subsequent Dirichlet tessellation of both the presence and integration points. B) Pseudo-random quadrature generated using random points from within the window. C) Grid quadrature where quadrature points are generated on a regular grid."}
library(ppmData)
path <- system.file("extdata", package = "ppmData")
lst <- list.files(path=path, pattern='*.tif',full.names = TRUE)
preds <- rast(lst)
presences <- subset(snails,SpeciesID %in% "Tasmaphena sinclairi")
npoints <- 1000
ppmdata1 <- ppmData(npoints = npoints, presences=presences,
                    window = preds[[1]], covariates=preds,
                    quad.method='quasi.random')
ppmdata2 <- ppmData(npoints = npoints, presences=presences,
                    window = preds[[1]], covariates=preds,
                    quad.method='pseudo.random')
ppmdata3 <- ppmData(npoints = npoints, presences=presences,
                    window = preds[[1]], covariates=preds,
                    quad.method='grid')

@@

There are a number of ways to setup the quadrature scheme to get an approximation of the integral $\Lambda = \int_{\bm{s}\in\mathcal{A}}\lambda(\bm{s})d\bm{s}$. One common, and arguably the easiest to program is to use a regular grid [@warton_poisson_2010]. This approach essentially creates a regular grid at a known resolution within a spatial window ($\mathcal{A} \subset R^2$). The grid approach is typically dense; and as we double the resolution it results in 4-times the number of quadrature points. The grid approach is not without detraction though. In particular, it is unlikely to handle edge effects well (when there is a consistent and potentially large pattern near the boundaries of $\mathcal{A}$). Then the grid will either be defined away from the edge, missing important trends, or defined on the edge, over-representing this trend. Also, another potential problem stems from the unlikely event that the spatial surface $\lambda(\bm{s}_i)$ has periodicity in its pattern aligned to the direction of the grid, such as repeated locations of mountain ridges. In that case, the grid-based quadrature will over- or under-estimate the integral depending on whether the grid is coincides with peaks or troughs of the surface.

@@

The package `spatstat` also allows a user to develop a quasi-random quadrature scheme, but this appears to be quite slow when used with Dirichlet weights for larger numbers of dummy (quadrature) points (\> 10000). For example, a regular window with 100000 quadrature points takes approximately 3.5 minutes to run in `spatstat` using `quadscheme` with `method=Dirichlet`. Our package computes the same Dirichlet tessellation in about 4.5 seconds. Typically, calculating a Dirichlet tessellation increases super-linearly and can be come slow with for large numbers of points. Part of the challenge when developing `ppmData` was to make this step efficient by using C++ code to run a Delaunay triangulation via a radial sweep algorithm [@sinclair_s-hull_2016]. We then calculate the Dirichlet tessellations as the dual graph of this triangulation. Finally, areal weights for quadrature are calculated as the area of each polygon in the Dirichlet tessellation.

@@

One of the main reasons for creating the `ppmData` package was to develop quadrature schemes for multiple species (marked) point process. The multiple species quadrature scheme can be used in multiple species models available for use in the `ecomix` package [@woolley_ecomix_2022], or models like joint species distribution models [@ovaskainen_making_2011]. Here is a quick example of how we would set up a quasi-random quadrature scheme for multiple species (Fig. 2). The main difference is the the column `speciesID` contains multiple species, `ppmData` recognises this internally and subsequently sets up a marked point process quadrature scheme. The quadrature scheme has a common set of quadrature locations, but the quadrature weights are calculated on a species specific basis $w_{ij}$, where $i$ is a sites index and $j$ is a species index.

@@

Here we demonstrate how to fit an IPPM using `glm` function and a `ppmData` object. This approach is very similar to the method presented in @warton_poisson_2010, except we are using a quasi-random quadrature. We use `glm` to fit an IPPM for simplicity, but if one wanted to fit an IPPM using different statistical machinery, then one could -- examples are: penalized regressions using `glmnet` [@friedman_regularization_2010] to achieve a `maxent` analysis [@renner_equivalence_2013], generalized additive models [e.g. @wood_generalized_2017] or statistical learning approaches [e.g. @hastie_elements_2009].

@@

We present example code for fitting an Inhomogeneous Poisson Process Model using `glm` from the R `stats` package [@r_core_team_r_2013]. We just need to extract the data.frame from the `ppmData` object and to specify a formula for the model. The only trick is making the response equal to `presence/weights` (the expectation). The right side of the formula will represent the covariates and their functional forms. In this example we specify independent 2$^{nd}$ degree polynomials for each covariate, except for the spatial coordinates X & Y which we include in the model as interacting 2$^{nd}$ degree polynomials. Once this is done, we define the weights in the `glm` function call and we now can fit an IPPM using `ppmData` and `glm`.

@@

```{r, fig.cap='Predicted intensity per for $\\approx $ 0.6 km$^2$ sized raster cell for the snail species \\textit{Tasmaphena sinclairi}. The points are the recoreded presences of \\textit{Tasmaphena sinclairi} in the region.', cache = TRUE, fig.align='center',fig.height=5,fig.width=5, echo=FALSE}
library(terra)
path <- system.file("extdata", package = "ppmData")
lst <- list.files(path=path, pattern='*.tif',full.names = TRUE)
preds <- rast(lst)
xy <- ppmFit::xyFromWindow(preds[[1]], mask.na = TRUE)
preds <- scale(preds)
preds2 <- c(xy,preds) 
pred <- predict(preds2, ft.ppm, type='response')
main <- ""#expression(paste("Predicted intensity per cell: ",lambda[i]))
p <- pred*prod(res(pred))
qtiles <- quantile(p[!is.na(p)],probs=seq(0,1,len=10),na.rm=TRUE)
breaks <- cut(p[!is.na(p)], breaks = qtiles,dig.lab = 2)

>>>>

Spatial data, the locations of objects in a spatial region, is a common type of data recorded in ecology, environmental sciences and epidemiology amongst many others. An appropriate statistical approach for such data are spatial point process models [@cressie_statistics_1993; @baddeley_practical_2000]. Spatial point processes can be used to understand the location of objects to inform decision making and improve ecological, health and other socio-economic outcomes. For example, the locations of individuals from a species' population are typically not distributed completely at random, but rather are observed at locations with favorable habitat conditions for that species. We might expect these species sightings are distributed according to relevant covariates.

Inhomogeneous Poisson Process Models allow for the spatial intensity of a point process to vary across space, usually as a function of spatially referenced covariates. One challenge when fitting an Inhomogeneous Poisson Process Model is the integral used to described the intensity surface in the log-likelihood (as described below) has no known closed analytic solution. For Inhomogeneous Poisson Process Models, the Berman-Turner device [@berman_approximating_1992] is commonly used to approximate the model using quadrature within a weighted Poisson generalized linear model. The type of quadrature scheme can have important ramifications for overall model fitting and inference [@warton_poisson_2010]. Approximating the integral accurately often comes at an increased computation cost, where a large number of quadrature points are required to robustly estimate the integral [@renner_point_2015].

@@

Here we present the `ppmData` R package that is designed to setup a quadrature scheme using Dirichlet tessellation for fitting spatial point process models. Our initial intention was to develop multiple species (marked) quadrature in order to improve computation of multiple species Poisson point process models. Since then, the package has evolved into one which can be useful for fitting univariate response inhomogeneous spatial point process, as discussed in this paper. We demonstrate how to set up quadrature for inhomogeneous Poisson Process Models and provide a simple example of how we can fit an inhomogeneous Poisson Process Model using a `ppmData` object.

@@

An Inhomogeneous Poisson Process Model is a statistical model used to describe a random point pattern for $n$ observed locations, ($\bm{y}=(y(\bm{s}_1),y(\bm{s_2}),\ldots,y(\bm{s}_n))$, in a spatial window $\mathcal{A}\subset R^2$. An Inhomogeneous Poisson Process Model assumes that the number of points in $\mathcal{A}$, $n$, is a Poisson random variable with total intensity $\Lambda$ defined as $\Lambda = \int_{\bm{s}\in\mathcal{A}}\lambda(\bm{s})d\bm{s}$. The spatial function $\lambda(\bm{s})$ is usually a non-constant function and defines the spatial surface that the spatial locations $\bm{y}$ are independently drawn from [@cressie_statistics_1993]. For convenience, we use subscript notation for particular spatial locations, for example, $\lambda_i=\lambda(\bm{s}_i)$ is the intensity surface of the point process at the $i^{th}$ location with coordinates $\bm{s}_i$. The intensity surface can be defined as $\log\{\lambda_i\}=\bm{x}_i^{\top}\bm{\beta}$ where $\bm{x}_i$ is a vector of spatially located covariates and $\bm{\beta}$ is a corresponding vector of coefficients.

@@

```{r fig1, fig.width=6, fig.height=3,fig.align='center',cache=TRUE, warning=FALSE, echo=FALSE, message=FALSE,fig.cap = "Quadrature schemes generated using the \\texttt{ppmData} package for the species \\textit{Tasmaphena sinclairi} located within Tasmania, Australia. The red points represent the known locations of \\textit{Tasmaphena sinclairi}. The black points represent the quadrature locations. A) Quasi-random quadrature where the integration points are generated using a quasi-random areal sample. Weights for each integration point are calculated as the dual of the subsequent Dirichlet tessellation of both the presence and integration points. B) Pseudo-random quadrature generated using random points from within the window. C) Grid quadrature where quadrature points are generated on a regular grid."}
library(ppmData)
path <- system.file("extdata", package = "ppmData")
lst <- list.files(path=path, pattern='*.tif',full.names = TRUE)
preds <- rast(lst)
presences <- subset(snails,SpeciesID %in% "Tasmaphena sinclairi")
npoints <- 1000
ppmdata1 <- ppmData(npoints = npoints, presences=presences,
                    window = preds[[1]], covariates=preds,
                    quad.method='quasi.random')
ppmdata2 <- ppmData(npoints = npoints, presences=presences,
                    window = preds[[1]], covariates=preds,
                    quad.method='pseudo.random')
ppmdata3 <- ppmData(npoints = npoints, presences=presences,
                    window = preds[[1]], covariates=preds,
                    quad.method='grid')

@@

There are a number of ways to setup a quadrature scheme to approximate the integral $\Lambda = \int_{\bm{s}\in\mathcal{A}}\lambda(\bm{s})d\bm{s}$. One common approach is to use a regular grid [@warton_poisson_2010], which creates a regular grid at a known resolution within a spatial window ($\mathcal{A} \subset R^2$). Despite the simplicity of the grid approach, it is not without detraction though. In particular, it is unlikely to handle edge effects well (when there is a consistent and potentially large pattern near the boundaries of $\mathcal{A}$). Then the grid will either be defined away from the edge, missing important trends, or defined on the edge, over-representing this trend. Also, another potential problem stems from the unlikely event that the spatial surface $\lambda(\bm{s}_i)$ has periodicity in its pattern aligned to the direction of the grid, such as repeated locations of mountain ridges. In that case, the grid-based quadrature will over- or under-estimate the integral depending on whether the grid is coincides with peaks or troughs of the surface.

@@

The package `spatstat` also allows a user to develop a quasi-random quadrature scheme, but this appears to be quite slow when used with Dirichlet weights for larger numbers of dummy (quadrature) points (\> 10000). For example, a regular window with 100000 quadrature points takes approximately 3.5 minutes to run in `spatstat` using `quadscheme` with `method=Dirichlet`. Our package computes the same Dirichlet tessellation in about 4.5 seconds. Typically, calculating a Dirichlet tessellation increases super-linearly and can become very slow for a large numbers of points. Part of the challenge when developing `ppmData` was to make this step efficient by a C++ implementation of a Delaunay triangulation radial sweep algorithm [@sinclair_s-hull_2016]. The Dirichlet tessellations are then calculated as the dual graph of the Delaunay triangulation. Areal weights for the quadrature points are calculated as the area of each Dirichlet tessellation polygon.

@@

One of the main reasons for creating the `ppmData` package was to develop quadrature schemes for multiple species (marked) point process. The multiple species quadrature scheme can be used in multiple species models available for use in the `ecomix` package [@woolley_ecomix_2022], or models like joint species distribution models [@ovaskainen_making_2011]. Here is a quick example of how we would set up a quasi-random quadrature scheme for multiple species (Fig. 2). The main difference is the column `speciesID` contains multiple species, `ppmData` recognises this internally and subsequently sets up a marked point process quadrature scheme. The quadrature scheme contains a common set of quadrature locations, but the quadrature weights are calculated on a species-specific basis $w_{ij}$, where $i$ is a site index and $j$ is a species index.

@@

Here we demonstrate how to fit an Inhomogeneous Poisson Process Model using `glm` function and a `ppmData` object. This approach is very similar to the method presented in @warton_poisson_2010, except we are using a quasi-random quadrature. We use `glm` to fit an Inhomogeneous Poisson Process Model for simplicity, but if one wanted to fit an Inhomogeneous Poisson Process Model using different statistical machinery, then one could -- examples are: penalized regressions using `glmnet` [@friedman_regularization_2010] to achieve a `maxent` analysis [@renner_equivalence_2013], generalized additive models [ @wood_generalized_2017] or statistical learning approaches [@hastie_elements_2009].

@@

We present a snippet of example code for fitting an Inhomogeneous Poisson Process Model using `glm` from the R `stats` package [@r_core_team_r_2013]. We just need to extract the data.frame from the `ppmData` object and to specify a formula for the model. The only trick is making the response equal to `presence/weights` (the expectation). The right side of the formula will represent the covariates and their functional forms. In this example we specify independent 2$^{nd}$ degree polynomials for each covariate, except for the spatial coordinates X & Y which we include in the model as interacting 2$^{nd}$ degree polynomials. Once this is done, we define the weights in the `glm` function call and we now can fit an IPPM using `ppmData` and `glm`.

@@

```{r, fig.cap='Predicted intensity per for $\\approx $ 0.6 km$^2$ sized raster cell for the snail species \\textit{Tasmaphena sinclairi}. The points are occurrence records for \\textit{Tasmaphena sinclairi} in the region.', cache = TRUE, fig.align='center',fig.height=5,fig.width=5, echo=FALSE}
library(terra)
path <- system.file("extdata", package = "ppmData")
lst <- list.files(path=path, pattern='*.tif',full.names = TRUE)
preds <- rast(lst)
xy <- ppmFit::xyFromWindow(preds[[1]], mask.na = TRUE)
preds <- scale(preds)
preds2 <- c(xy,preds) 
pred <- predict(preds2, ft.ppm, type='response')
main <- ""#expression(paste("Predicted intensity per cell: ",lambda[i]))
p <- pred*prod(res(pred))
qtiles <- quantile(p[!is.na(p)],probs=seq(0,1,len=10),na.rm=TRUE)
breaks <- cut(p[!is.na(p)], breaks = qtiles,dig.lab = 2)
